---
title: "Streamlining STEMI Care: Quality of Care Report for ST-Elevation Myocardial Infarction (STEMI) Patients in SyntheticMGUH"
description: "Quality improvement analysis of door-to-balloon time in a synthetic hospital EHR"
author:
  - name: "Jay Nagabhairu"
  - name: "Stephanie Araki"
  - name: "Maxwell Lewis"
  - name: "Pari Shah"
date: last-modified
date-format: short
title-block-banner: '#041E42'
format:
  html:
    self-contained: true
---

```{r setup, include=FALSE}
#| warning: false
library(tidyverse)     # loads the tidyverse tools
library(RPostgres)     # loads the database driver for PostgreSQL
library(connections)   # helps RPostgres work with RStudio
library(keyring)       # access to a local encrypted keychain for passwords


con <- connection_open(RPostgres::Postgres(),
          dbname = "syntheticmguh",
          host = "34.145.215.95",
          user = "hids502_student",
          password = key_get(service = "syntheticmguh", 
                             username = "hids502_student"),
          # Tell the driver to return very large integers as floating point (vs truncating them)
          bigint = "numeric")

```

# I. Introduction

ST-Elevation Myocardial Infarction (STEMI) is a severe type of heart attack caused by complete blockage of a coronary artery. It is a medical emergency requiring timely diagnosis and treatment to restore blood flow and reduce myocardial damage. Evidence-based guidelines emphasize rapid identification and management of STEMI patients to optimize outcomes and reduce mortality.

Quality Improvement (QI) initiatives play a crucial role in evaluating and enhancing adherence to these guidelines. This report evaluates key metrics related to STEMI care utilizing patient data from the [Synthea dataset](https://github.com/synthetichealth/synthea/wiki/CSV-File-Data-Dictionary) to assess compliance with clinical standards, identify areas for improvement, and propose strategies to enhance care quality.

Key metrics for compliance, derived from the [American Heart Association (AHA)/American College of Cardiology (ACC) STEMI guidelines](https://www.ahajournals.org/doi/10.1161/cir.0b013e3182742cf6), include:

* **Door-to-Balloon (DTB) time**: The proportion of patients receiving primary percutaneous coronary intervention (PCI) within the recommended 90-minute period.
* **Medication Compliance**: The proportion of patients prescribed dual antiplatelet therapy (DAPT), high-intensity statins, and beta-blockers within 24 hours of a STEMI encounter.
* **3-month Mortality Rate**: The proportion of patients with reported death within 3 months of a STEMI encounter.

# II. Establishing a Myocardial Infarction (MI) Cohort

To ensure our analysis encompasses a comprehensive range of patients with STEMI, we first created a general cohort of patients who presented with myocardial infarction (MI) based on the following criteria:

* An **encounter** for myocardial infarction, from the `encounters` table.
* A **problem/condition** of myocardial infarction, from the `conditions` table.
* A **procedure** for myocardial infarction, from the `procedures` table.

| SNOMED-CT Code | Description |
| ---------------|-------------|
| 22298006  | Myocardial infarction |
| 401303003	| Acute ST segment elevation myocardial infarction (disorder) |
| 401314000	| Acute non-ST segment elevation myocardial infarction (disorder)	|

* Note: We purposefully excluded patients with a History of myocardial infarction (399211009) in attempts to capture only patients who received a STEMI diagnosis from our hospital.


The following demographic data points from the `patients` table will also be included:

| Demographic | Description |
| ---------------|-------------|
| `patient` | Patient's unique identifier. |
| `birthdate` | Patient's date of birth. |
| `deathdate`	| Patient's date of death. If the patient has no date of death (either alive or unknown), NULL. |
| `gender` | Patient's gender. M is male and F is female. |
| `race` | Patient's race. |
| `ethnicity` | Patient's ethnicity. |
| `zip` | Patient's zip code. |

```{sql connection=con}
-- Drop the temporary table if it already exists
DROP TABLE IF EXISTS mi_cohort;
```

```{sql connection=con}
CREATE TEMPORARY TABLE mi_cohort AS
WITH all_mi_patients AS (
    -- MI encounter
    SELECT 
        patient,
        reasoncode AS diagnosis_code,
        reasondescription AS diagnosis_desc,
        start
    FROM encounters
    WHERE reasoncode IN ('22298006', '401303003', '401314000') -- Myocardial infarction codes

    UNION

    -- MI condition
    SELECT 
        patient,
        code AS diagnosis_code,
        description AS diagnosis_desc,
        start
    FROM conditions
    WHERE code IN ('22298006', '401303003', '401314000') -- Myocardial infarction codes

    UNION

    -- MI procedure
    SELECT 
        patient,
        reasoncode AS diagnosis_code,
        reasondescription AS diagnosis_desc,
        start
    FROM procedures
    WHERE reasoncode IN ('22298006', '401303003', '401314000') -- Myocardial infarction codes
),
-- if a patient has multiple diagnoses, use only the most recent diagnosis
recent_diagnosis AS (
    SELECT 
        patient,
        diagnosis_code,
        diagnosis_desc,
        MAX(start) AS most_recent_date
    FROM all_mi_patients
    GROUP BY patient, diagnosis_code, diagnosis_desc
),
most_recent_per_patient AS (
    SELECT DISTINCT ON (patient) 
        patient, 
        diagnosis_code, 
        diagnosis_desc,
        most_recent_date
    FROM recent_diagnosis
    ORDER BY patient, most_recent_date DESC
)
SELECT 
    m.patient,
    m.diagnosis_code,
    m.diagnosis_desc,
    m.most_recent_date,
    p.birthdate, 
    p.deathdate, 
    p.race, 
    p.ethnicity, 
    p.gender,
    p.city,
    p.state,
    p.zip
FROM most_recent_per_patient m
JOIN patients p 
    ON p.id = m.patient;
```

```{sql connection=con}
 -- Report the total number of patients in the MI cohort
SELECT COUNT(DISTINCT patient) AS total_mi_patients
FROM mi_cohort
;
```
# III. Establishing the ST-Elevation Myocardial Infarction (STEMI) Cohort
From the broader MI cohort, we then assembled a cohort of STEMI patients based on the following criteria:

* An ECG/EKG procedure
* A cardiology consultation 
* A PCI procedure

```{sql connection=con}
-- Drop the temporary table if it already exists
DROP TABLE IF EXISTS stemi_cohort;
```

```{sql connection=con}
CREATE TEMPORARY TABLE stemi_cohort AS

WITH ekg_patients AS (
  SELECT 
    patient
  FROM procedures
  -- EKG-related procedure codes:
  WHERE code = '29303009'  -- Electrocardiographic procedure (procedure)
     OR code = '268400002'  -- 12 lead electrocardiogram (procedure)
)

, cardio_consult_patients AS (
  SELECT
    patient
  FROM procedures
    WHERE code = '698314001' -- Consultation for treatment (procedure)
)

, pci_patients AS (
  SELECT 
    patient
  FROM procedures
  WHERE code = '415070008'  -- Percutaneous coronary intervention (procedure)
)

SELECT 
  DISTINCT pci.patient,
  p.birthdate, 
  p.deathdate, 
  p.race, 
  p.ethnicity, 
  p.gender,
  p.city,
  p.state,
  p.zip
FROM pci_patients pci
JOIN patients p ON pci.patient = p.id
JOIN ekg_patients ekg ON pci.patient = ekg.patient
JOIN cardio_consult_patients cons ON pci.patient = cons.patient
WHERE pci.patient IN (SELECT patient FROM mi_cohort);
```

```{sql connection=con}
-- Check that all stemi_cohort patients are also in mi_cohort
SELECT COUNT(*)
FROM stemi_cohort s
WHERE s.patient NOT IN (SELECT patient FROM mi_cohort);
-- should return 0 
```

```{sql connection=con}
-- Check that each row in the stemi_cohort table corresponds to a unique patient; these values should be equivalent
SELECT COUNT(patient) AS all_patients,
COUNT(DISTINCT patient) AS distinct_patients
FROM stemi_cohort
;
```
Approximately 26% of the MI cohort was included in the STEMI cohort.

```{sql connection=con}
-- Calculate the percentage of MI patients who are in the STEMI cohort
SELECT
  (SELECT COUNT(DISTINCT patient) FROM mi_cohort) AS mi_cohort_count,
  (SELECT COUNT(DISTINCT patient) FROM stemi_cohort) AS stemi_cohort_count,
  ROUND((COUNT(DISTINCT stemi_cohort.patient) * 100.0 / (SELECT COUNT(DISTINCT mi_cohort.patient) FROM mi_cohort)), 2) AS percent_STEMI_from_MI
FROM stemi_cohort
;
```

## A. STEMI Cohort - Demographics
The demographic characteristics of the STEMI cohort provide insight into the population being analyzed. Key findings include:

### 1. Age
Here, we generate general statistics for the age of the STEMI cohort and also categorize patients into age ranges, where each range is inclusive (e.g. 'BETWEEN 36 AND 55' includes 36 and 55). 
The age distribution shows a majority of the patients are between the ages of 66 and 89.

#### Age Statistics:
```{sql connection=con}
-- Calculate the average, minimum and maximum age of patients in the STEMI cohort
SELECT
  AVG(date_part('year', age('2023-12-31', birthdate))) AS avg_age,
  MIN(date_part('year', age('2023-12-31', birthdate))) AS min_age,
  MAX(date_part('year', age('2023-12-31', birthdate))) AS max_age
FROM stemi_cohort
;
```

#### Age Distribution by Bins:
```{sql connection=con}
-- Calculate the proportion (as a percentage) of patients in each age category
SELECT
    CASE 
      WHEN date_part('year', age('2023-12-31', birthdate)) < 18 THEN '< 18'
      WHEN date_part('year', age('2023-12-31', birthdate)) BETWEEN 18 AND 35 THEN '18-35'
      WHEN date_part('year', age('2023-12-31', birthdate)) BETWEEN 36 AND 55 THEN '35-55'
      WHEN date_part('year', age('2023-12-31', birthdate)) BETWEEN 56 AND 65 THEN '56-65'
      WHEN date_part('year', age('2023-12-31', birthdate)) BETWEEN 66 AND 89 THEN '66-89'
      WHEN date_part('year', age('2023-12-31', birthdate)) > 89 THEN '> 89'
      ELSE 'Unknown Age Group'
    END AS age_group,
    COUNT(*) as patient_count,
    ROUND((COUNT(*) * 100.0 / SUM(COUNT(*)) OVER ()), 2) AS proportion_percent
FROM stemi_cohort
GROUP BY age_group
ORDER BY age_group ASC
;
```

### 2. Gender
The gender distribution shows a predominance of males over females. 

```{sql connection=con}
SELECT gender, 
  COUNT(*) AS num_patients,
  ROUND((COUNT(*) * 100.0 / SUM(COUNT(*)) OVER ()), 2) AS proportion_percent
FROM stemi_cohort
GROUP BY gender
;
```

### 3. Race and Ethnicity
The majority of the patients in the STEMI cohort identify as either black or white, and non-hispanic. 

```{sql connection=con}
SELECT race, 
  COUNT(*) AS num_patients,
  ROUND((COUNT(*) * 100.0 / SUM(COUNT(*)) OVER ()), 2) AS proportion_percent
FROM stemi_cohort
GROUP BY race
ORDER BY num_patients DESC
;
```
```{sql connection=con}
SELECT ethnicity, 
  COUNT(*) AS num_patients,
  ROUND((COUNT(*) * 100.0 / SUM(COUNT(*)) OVER ()), 2) AS proportion_percent
FROM stemi_cohort
GROUP BY ethnicity
ORDER BY num_patients DESC
;
```

## B. STEMI Cohort - Risk Factors / Comorbidities

### 1. Smoking Status
> What proportion of patients in the STEMI cohort reported formerly or currently smoking?

The smoking status of patients was analyzed to consider the potential impact of tobacco use on MI risk. 

The majority of patients report never having smoked, and 33% report formerly smoking.

```{sql connection=con}
-- Calculate the proportion of patients in the stemi_cohort who have or currently smoked
WITH smokers AS (
  SELECT
      patient,
      CASE 
        WHEN code = '72166-2' AND value = 'Smokes tobacco daily (finding)' THEN 'Current smoker'
        WHEN code = '72166-2' AND value = 'Ex-smoker (finding)' THEN 'Former smoker'
        WHEN code = '72166-2' AND value = 'Never smoked tobacco (finding)' THEN 'Never smoked'
      END AS smoking_status,
      ROW_NUMBER() OVER (PARTITION BY patient ORDER BY date DESC) AS rn
  FROM observations
  WHERE code = '72166-2'
)
SELECT
  smoking_status,
  COUNT(DISTINCT sm.patient) AS patient_count,
  ROUND((COUNT(DISTINCT sm.patient) * 100.0 / (SELECT COUNT(DISTINCT patient) FROM stemi_cohort)), 2) AS proportion_percent
FROM smokers sm
RIGHT JOIN stemi_cohort st ON sm.patient = st.patient
WHERE sm.rn = 1 -- Only take the most recent record per patient
GROUP BY smoking_status
;
```

### 2. Hypertension
> What proportion of patients in the STEMI cohort have a reported condition of hypertension? 

Hypertension is a significant risk factor for STEMI due to its role in the pathophysiology of coronary artery disease (CAD), which is the underlying cause of STEMI.

The majority of the STEMI cohort report a condition of hypertension (77%).

```{sql connection=con}
-- Calculate the proportion of patients in the stemi_cohort who have a condition of hypertension (htn)
WITH htn_patients AS (
  SELECT patient
  FROM conditions
  WHERE code = '59621000' -- Essential hypertension (disorder)
)

SELECT 
  COUNT(DISTINCT s.patient) AS htn_patient_count,
  (SELECT COUNT(DISTINCT patient) FROM stemi_cohort) AS total_patient_count,
  ROUND((COUNT(DISTINCT s.patient) * 100.0 / (SELECT COUNT(DISTINCT patient) FROM stemi_cohort)), 2) AS htn_percent
FROM stemi_cohort s
JOIN htn_patients h ON s.patient = h.patient;
```

# IV. Quality Measurements

## A. Door-to-Balloon (DTB) Time
> What is the distribution of DTB times for STEMI patients?

A simplified STEMI clinical workflow includes the following steps:
```{mermaid}
flowchart LR
  A(ED Admission) --> B(EKG) 
  B --> C(Cardiology Consultation) 
  C --> D(PCI)
```

The time between a patient’s arrival in the Emergency Department (ED) and the beginning of percutaneous coronary intervention (PCI) is referred to as **door-to-balloon time (DTB time)**. Clinical guidelines emphasize that DTB time should be under 90 minutes to minimize the risk of death and long-term complications. 

Here, we assume that the encounter begins at the start of the EKG procedure, and, therefore, calculate the **DTB time as the difference between the start of the EKG procedure and the start of the PCI procedure**. 

Using the cohort of STEMI patients we previously established, we assembled a timeline of events for each patient, reflecting the duration of and between each step in the clinical workflow, and the total DTB time. 

While the STEMI cohort includes over 400 patients, the STEMI timeline table only includes patients who have complete records of the three crucial steps in the clinical workflow: EKG, Cardiology Consultation, and PCI. Additionally, some STEMI patients receive surgical intervention instead of PCI, so those patients are also not reflected in the timeline. 

```{sql connection=con}
-- Drop the temporary table if it already exists
DROP TABLE IF EXISTS stemi_timeline;
```

```{sql connection=con}
CREATE TEMPORARY TABLE stemi_timeline AS

WITH ekg_patients AS (
  SELECT 
    patient,
    encounter AS ekg_encounter,
    start AS ekg_start,
    stop AS ekg_stop,
    code AS ekg_code,
    description AS ekg_desc,
    EXTRACT(YEAR FROM start) AS ekg_year,  -- Extract the year from the EKG start time
    ROW_NUMBER() OVER (PARTITION BY patient ORDER BY start DESC) AS ekg_rank  -- Pick the most recent EKG
  FROM procedures
  -- EKG-related procedure codes:
  WHERE code = '29303009'  -- Electrocardiographic procedure (procedure)
     OR code = '268400002'  -- 12 lead electrocardiogram (procedure)
)

, cardio_consult_patients AS (
  SELECT
    patient,
    encounter AS consult_encounter,
    start AS consult_start,
    stop AS consult_stop,
    code AS consult_code,
    description AS consult_desc,
    EXTRACT(YEAR FROM start) AS consult_year,  -- Extract the year from the Consult start time
    ROW_NUMBER() OVER (PARTITION BY patient ORDER BY start DESC) AS consult_rank  -- Pick the most recent consult
  FROM procedures
  WHERE code = '698314001' -- Consultation for treatment (procedure)
)

, pci_patients AS (
  SELECT 
    patient,
    encounter AS pci_encounter,
    start AS pci_start,
    stop AS pci_stop,
    EXTRACT(YEAR FROM start) AS pci_year,  -- Extract the year from the PCI start time
    ROW_NUMBER() OVER (PARTITION BY patient ORDER BY start DESC) AS pci_rank  -- Pick the most recent PCI
  FROM procedures
  WHERE code = '415070008'  -- Percutaneous coronary intervention (procedure)
)

SELECT 
  s.patient, 
  -- Encounter IDs
  e.ekg_encounter, 
  c.consult_encounter,
  p.pci_encounter,
  -- Datetimes
  e.ekg_start,
  e.ekg_stop,
  c.consult_start,
  c.consult_stop,
  p.pci_start,
  p.pci_stop,
  -- Calculate durations:
  -- EKG
  ROUND(EXTRACT(EPOCH FROM (e.ekg_stop - e.ekg_start)) / 60, 2) AS ekg_duration_minutes,  
  ROUND(EXTRACT(EPOCH FROM (c.consult_start - e.ekg_stop)) / 60, 2) AS ekg_to_consult_minutes,
  -- Consult
  ROUND(EXTRACT(EPOCH FROM (c.consult_stop - c.consult_start)) / 60, 2) AS consult_duration_minutes,
  ROUND(EXTRACT(EPOCH FROM (p.pci_start - c.consult_stop)) / 60, 2) AS consult_to_pci_minutes,
  -- PCI
  ROUND(EXTRACT(EPOCH FROM (p.pci_stop - p.pci_start)) / 60, 2) AS pci_duration_minutes,
  -- DTB
  ROUND(EXTRACT(EPOCH FROM (p.pci_start - e.ekg_start)) / 60, 2) AS dtb_minutes 
FROM stemi_cohort s
JOIN ekg_patients e ON s.patient = e.patient
JOIN cardio_consult_patients c ON s.patient = c.patient
JOIN pci_patients p ON s.patient = p.patient
-- FILTERS
-- Add column to reflect if the encounters for ekg, consult and pci are the same or different
WHERE p.pci_rank = 1  -- Only the most recent PCI procedure
  AND e.ekg_rank = 1  -- Only the most recent EKG procedure
  AND c.consult_rank = 1  -- Only the most recent consult
  -- Filter out negative durations:
  AND EXTRACT(EPOCH FROM (e.ekg_stop - e.ekg_start)) >= 0  -- Filter out rows with negative EKG durations
  AND EXTRACT(EPOCH FROM (c.consult_stop - c.consult_start)) >= 0  -- Filter out rows with negative consult durations
  AND EXTRACT(EPOCH FROM (p.pci_stop - p.pci_start)) >= 0  -- Filter out rows with negative PCI durations
  AND EXTRACT(EPOCH FROM (p.pci_start - e.ekg_start)) >= 0  -- Filter out rows with negative DTB times
  -- Ensure all encounters are from the same year (same visit):
  AND e.ekg_year = c.consult_year
  AND c.consult_year = p.pci_year
  -- Ensure encounters occur within a reasonable time window (within 24 hours):
  AND EXTRACT(EPOCH FROM (p.pci_start - e.ekg_start)) <= 86400  -- 24 hours (86400 seconds)
ORDER BY p.patient;
```

```{sql connection=con}
-- sanity check: these values should match
SELECT COUNT(DISTINCT patient) AS distinct_count,
COUNT(patient) AS all_patients
FROM stemi_timeline
```
```{sql connection=con}
-- STEMI patients in the stemi_timeline table should have been admitted to the ED
SELECT 
  COUNT (DISTINCT s.patient) AS patient_count,
  e.encounterclass, 
  e.description
FROM stemi_timeline s
JOIN encounters e ON s.patient = e.patient AND s.ekg_encounter = e.id
WHERE encounterclass = 'emergency'
GROUP BY e.encounterclass, e.description
```

Export stemi_timeline_pre_intervention.csv for Tableau analysis:
```{r}
query1 <- "SELECT * FROM stemi_timeline"
stemi_timeline_pre_intervention <- dbGetQuery(con, query1)

# Get the current timestamp
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")

# Create a filename with the timestamp
filename <- paste0("stemi_timeline_pre_intervention_", timestamp, ".csv")

# Write the data to a CSV file
write.csv(stemi_timeline_pre_intervention, filename, row.names = FALSE)
```

#### DTB time distribution in the STEMI cohort
95% of STEMI patients were transferred to the cath lab for PCI within 30 minutes of arrival. Approximately 4% of patients experienced delayed treatment with DTB times over 90 minutes, exceeding clinical guidelines.
```{sql connection=con}
SELECT 
  CASE 
    WHEN dtb_minutes < 30 THEN '<30 minutes'
    WHEN dtb_minutes BETWEEN 31 and 60 THEN '31-60 minutes'
    WHEN dtb_minutes BETWEEN 61 and 90 THEN '61-90 minutes'
    WHEN dtb_minutes > 90 THEN '90 minutes'
    ELSE 'Unknown'
  END AS dtb_bin,
  COUNT(patient) AS patient_count,
  ROUND((COUNT(patient) * 100.0 / SUM(COUNT(patient)) OVER ()), 2) AS proportion_percent
FROM stemi_timeline
GROUP BY dtb_bin
ORDER BY dtb_bin
```

```{sql connection=con, output.var="dtb_dist_df"}
SELECT 
  CASE 
    WHEN dtb_minutes < 30 THEN '1. <30 minutes'
    WHEN dtb_minutes BETWEEN 31 and 60 THEN '2. 31-60 minutes'
    WHEN dtb_minutes BETWEEN 61 and 90 THEN '3. 61-90 minutes'
    WHEN dtb_minutes > 90 THEN '4.>90 minutes'
    ELSE 'Unknown'
  END AS dtb_bin,
  COUNT(patient) AS patient_count,
  ROUND((COUNT(patient) * 100.0 / SUM(COUNT(patient)) OVER ()), 2) AS proportion_percent
FROM stemi_timeline
GROUP BY dtb_bin
ORDER BY dtb_bin
```
```{r}
# plot the distribution of deaths by time
ggplot(dtb_dist_df, aes(x = dtb_bin, y = patient_count)) +
  geom_col(fill = "#041E42", color = "black") +
  labs(title = "Distribution of DTB Times for STEMI Patients",
       x = "DTB Time",
       y = "Patient Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

For the eight patients who did not receive PCI within the 90 minute window, one patient had a reported death within 3 months of the encounter. Therefore, it is difficult to conclude if there is a relationship between 3-month mortality and longer DTB times from this data. A larger sample size could provide further insight into a potential correlation. 

```{sql connection=con}
SELECT 
  patient,
  ekg_start,
  dtb_minutes,
  deathdate,
  CASE 
    WHEN (deathdate - ekg_start) <= INTERVAL '3 months' THEN 1
    ELSE 0
  END AS stemi_death
FROM stemi_timeline
JOIN patients ON stemi_timeline.patient = patients.id
WHERE dtb_minutes >= 90;
```

## B. Medication Compliance
> Whhat proportion of patients in the stemi_timeline table received medication within 24hrs of their visit?

For STEMI patients, receiving the appropriate medications within 24 hours of their visit is crucial for several reasons:

**Dual Antiplatelet Therapy (DAPT):** This involves using medications like aspirin and another antiplatelet agent (prasugrel, ticagrelor, or clopidogrel). DAPT helps prevent the formation of clots, which are a primary cause of STEMI. These medications inhibit platelet aggregation and reduce the risk of recurrent myocardial infarction and other thrombotic events by stabilizing the coronary artery after it has been opened during the acute phase.

**High-Intensity Statins:** Statins are used to lower LDL cholesterol and reduce the risk of further coronary events by stabilizing atherosclerotic plaques and reducing inflammation. High-intensity statins are particularly beneficial in STEMI patients as they help to minimize the risk of future cardiovascular events.

**Beta-Blockers:** These medications help reduce the heart's workload and oxygen demand, which can improve heart function post-STEMI. They slow the heart rate, reduce blood pressure, and diminish the force of contraction of the heart muscle, thereby lowering the risk of further damage to the heart tissue.

Medications categorized by type:

| Medication Type | Medications |
|-----------------|---------------|
| Dual Antiplatelet Therapy (DAPT) | Aspirin, Prasugrel, Ticagrelor, Clopidogrel, Eptifibatide, Tirofiban, Bivalirudin |
| High-Intensity Statins | Enoxaparin |
| Beta-Blockers | Heparin, Fondaparinux |

Clinical guidelines strongly recommend the administration of these medications to STEMI patients as a standard part of the acute management protocol. Adherence to these guidelines is associated with improved patient outcomes, reinforcing the importance of medication compliance within the critical first 24 hours.

Here, we see that 99% of STEMI patients received timely medication. 

```{sql connection=con}
WITH stemi_meds AS (
SELECT
  s.patient,
  -- DAPTs
  MAX(CASE 
    WHEN m.code IN (2563431, 243670, 855812, 212033, 1116635, 749196, 1997015, 200349, 1737466, 309362) THEN 1
    ELSE 0
  END) AS DAPT,
  -- Statins  
  MAX(CASE 
    WHEN m.code IN (854255) THEN 1
    ELSE 0
  END) AS Statin,
  -- BetaBlockers
  MAX(CASE 
    WHEN m.code IN (1659263, 1361048, 1361226, 861356) THEN 1
    ELSE 0
  END) AS BetaBlocker,
  -- Other
  MAX(CASE 
    WHEN m.code NOT IN (2563431, 243670, 855812, 212033, 1116635, 749196, 1997015, 200349, 1737466, 309362, 854255, 1659263, 1361048, 1361226, 861356) THEN 1
    ELSE 0
  END) AS Other
FROM medications AS m
JOIN stemi_timeline AS s ON m.patient = s.patient AND m.encounter = s.ekg_encounter
WHERE
  -- Filter for medications administered within 24hrs of STEMI encounter
  (EXTRACT(EPOCH FROM (m.start - s.ekg_start)) / 60) <= 1440 -- 24 hours = 1440 minutes
  AND s.patient IN (SELECT patient FROM stemi_timeline)
GROUP BY s.patient
ORDER BY s.patient
)

SELECT
  -- SUM(DAPT) AS DAPT,
  ROUND((SUM(DAPT) / SUM(COUNT(patient)) OVER ()), 2) AS dapt_percent,
  -- SUM(Statin) AS Statin,
  ROUND((SUM(Statin) / SUM(COUNT(patient)) OVER ()), 2) AS statin_percent,
  -- SUM(BetaBlocker) AS BetaBlocker,
  ROUND((SUM(BetaBlocker) / SUM(COUNT(patient)) OVER ()), 2) AS betablocker_percent,
  -- SUM(Other) AS Other,
  ROUND((SUM(Other) / SUM(COUNT(patient)) OVER ()), 2) AS other_percent
FROM stemi_meds;
```

## C. 3-Month Mortality Rate
> What proportion of patients in the stemi_timeline table had a STEMI-related death? 

The 3-month mortality rate is a critical early indicator of the effectiveness of acute STEMI management and treatment strategies. It helps in assessing the immediate outcomes of the care provided to patients following a STEMI. A high mortality rate within the first three months could signal issues with the initial intervention or post-discharge management, necessitating review and potential changes to clinical protocols.

Here, we define STEMI-related deaths as patient deaths that occur within 3 months of a STEMI encounter. Approximately 13% of the patients in the STEMI timeline table had reported death dates, and of those, 7% of deaths occurred within 3 months. 

```{sql connection=con}
DROP TABLE IF EXISTS stemi_deaths;
```

```{sql connection=con}
CREATE TEMPORARY TABLE stemi_deaths AS
  SELECT 
  s.patient,
  s.ekg_encounter,
  s.ekg_start,
  p.deathdate,
  CASE 
  WHEN p.deathdate IS NOT NULL 
    AND p.deathdate <= s.ekg_start + INTERVAL '3 months' 
  THEN 1 
  ELSE 0 
END AS stemi_death
FROM stemi_timeline s
JOIN patients p ON s.patient = p.id;
```

```{sql connection=con}
-- Calculate the number of deaths for patients in the STEMI timeline table
SELECT 
  COUNT(DISTINCT patient) AS total_stemi_patients,
  COUNT(p.deathdate) AS total_deaths,
  ROUND(COUNT(p.deathdate) * 100.0 / COUNT(DISTINCT s.patient), 2) AS percent_total_deaths,
  SUM(CASE 
        WHEN p.deathdate IS NOT NULL 
          AND p.deathdate <= s.ekg_start + INTERVAL '3 months' 
        THEN 1 
        ELSE 0 
      END) AS deaths_within_3_months,
  ROUND(SUM(CASE 
        WHEN p.deathdate IS NOT NULL 
          AND p.deathdate <= s.ekg_start + INTERVAL '3 months' 
        THEN 1 
        ELSE 0 
      END) * 100.0 / COUNT(DISTINCT patient), 2) AS percent_deaths_within_3_months
FROM stemi_timeline s
JOIN patients p ON s.patient = p.id;
```

```{sql connection=con, output.var = "death_dist_df",}
-- Plot the distribution of STEMI deaths by time 
SELECT 
  CASE 
    WHEN deathdate - ekg_start <= INTERVAL '3 months' THEN '1. < 3 months'
    WHEN deathdate - ekg_start > INTERVAL '3 months' AND deathdate - ekg_start <= INTERVAL '6 months' THEN '2. 3-6 months'
    WHEN deathdate - ekg_start > INTERVAL '6 months' AND deathdate - ekg_start <= INTERVAL '1 year' THEN '3. 6 months-1 year'
    WHEN deathdate - ekg_start > INTERVAL '1 year' AND deathdate - ekg_start <= INTERVAL '3 years' THEN '4. 1-3 years'
    WHEN deathdate - ekg_start > INTERVAL '3 years' AND deathdate - ekg_start <= INTERVAL '5 years' THEN '5. 3-5 years'
    ELSE '6. > 5 years'
  END AS time_stemi_to_death,
  COUNT(*) AS patient_count
FROM stemi_deaths
WHERE deathdate IS NOT NULL
GROUP BY time_stemi_to_death
-- ORDER BY time_stemi_to_death
;
```
```{r}
# plot the distribution of deaths by time
ggplot(death_dist_df, aes(x = time_stemi_to_death, y = patient_count)) +
  geom_col(fill = "#041E42", color = "black") +
  labs(title = "Distribution of STEMI Patient Deaths",
       x = "Time from STEMI Encounter to Death",
       y = "Patient Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Export stemi_deaths_pre_intervention.csv for Tableau analysis:
```{r}
query2 <- "SELECT * FROM stemi_deaths"
stemi_deaths_pre_intervention <- dbGetQuery(con, query2)

# Get the current timestamp
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")

# Create a filename with the timestamp
filename <- paste0("stemi_deaths_pre_intervention_", timestamp, ".csv")

# Write the data to a CSV file
write.csv(stemi_deaths_pre_intervention, filename, row.names = FALSE)
```

## D. Troponin Laboratory Tests
> What proportion of patients in the stemi_timeline table received a troponin laboratory test during their STEMI encounter?

Troponin is a protein released into the bloodstream when heart muscle cells are damaged, making it a sensitive and specific marker for myocardial infarction. Elevated troponin levels within 3 to 6 hours of symptom onset are a key diagnostic criterion for STEMI. Measuring troponin helps distinguish between non-ST-elevation myocardial infarction (NSTEMI) and STEMI, guiding appropriate treatment plans.

100% of STEMI patients received a troponin test during their STEMI encounter.

```{sql connection=con}
WITH trop_labs AS (
  SELECT patient, encounter, code, description
  FROM observations
  WHERE code = '89579-7' -- Troponin I.cardiac [Mass/volume] in Serum or Plasma by High sensitivity method	
)

-- SELECT COUNT(DISTINCT s.patient)
SELECT 
  COUNT(s.patient) AS troponin_patient_count,
  ROUND((COUNT(s.patient) * 100.0 / SUM(COUNT(s.patient)) OVER ()), 2) AS troponin_percent
FROM stemi_timeline s
JOIN trop_labs t ON s.patient = t.patient AND s.ekg_encounter = t.encounter
```

